---
title: "Standaridized Boolean Regulatory Function Bias"
author: "[John Zobolas](https://github.com/bblodfon)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
description: "Data analyses related to Truth Density bias in standardized boolean regulatory functions"
url: 'https\://bblodfon.github.io/brf-bias/'
github-repo: "bblodfon/brf-bias"
bibliography: references.bib
link-citations: true
site: bookdown::bookdown_site
---

# Input Libraries {-}

Libraries used:
```{r Load libraries, message = FALSE}
library(gtools)
library(dplyr)
library(tibble)
library(tidyr)
library(usefun)
library(foreach)
library(doParallel)
library(DT)
library(ggplot2)
library(ggpubr)
library(corrplot)
library(latex2exp)
```

## Link-operator Boolean Regulatory Functions {-}

Let $f$ be a boolean function $f(x,y):\{0,1\}^n \rightarrow \{0,1\}$, with $m \ge 1$ **activators** $x=\{x_i\}_{i=1}^{m}$ and $k \ge 1$ **inhibitors** $y=\{y_i\}_{i=1}^{k}$, that is a total of $n = m + k$ regulators.
The link-operator boolean functions have a (non-DNF) formula representation that partitions the two distinct type of regulators (positive regulators are called *activators*, negative regulators are called *inhibitors*) into two separate groups and connect them with a specific logical operator (which we call the *link operator*).
The presence of the link operator is what forces the condition $m,k \ge 1$ (at least one regulator in each category).
An example of such a function that has been used in the literature is the formula with the `AND-NOT` link operator [@Mendoza2006]:

- `AND-NOT`: $$f_{AND-NOT}(x,y) = \left(\bigvee_{i=1}^{m} x_i\right) \land \lnot \left(\bigvee_{i=1}^{k} y_i\right)$$

A **variation** of this one is the `OR-NOT` link operator function:

- `OR-NOT`: $$f_{OR-NOT}(x,y) = \left(\bigvee_{i=1}^{m} x_i\right) \lor \lnot \left(\bigvee_{i=1}^{k} y_i\right)$$

In theory, we could use any of the known logical operators as a link-operator, e.g. `XOR`, `NAND`, simple `AND` etc.

For example, another link-operator function is the next one:

- `BalanceOp1`: $$f(x,y) = \bigvee_{\forall (i,j)}^{m,k}(x_i\land \lnot y_j) = \left(\bigvee_{i=1}^{m} x_i\right) \land \left(\bigvee_{i=1}^{k} \lnot y_i\right)$$

:::{.note}
@Cury2019 defined the **consistent boolean regulatory** functions and their respective **complete DNF forms (CDNF)**.

The link-operator functions listed above are a subset of these functions, since they respect the regulatory structure in their respective DNF forms (for the *BalanceOp1* it's evident since it's written first in the DNF - for the *AND-NOT* and *OR-NOT* functions, see their respective DNF forms in the truth density proofs \@ref(prp:and-not-proof) and \@ref(prp:or-not-proof).
:::

## Threshold functions {-}

The **boolean threshold functions** are formulated based on given *pseudo-Boolean* constrains for the regulators.
Thus, the activators and inhibitors are combined in an *additive* manner and not a *combinatorial* one and the **majority rule** defines the value of the function [@Chaouiya2013].

Two simple cases of **threshold** functions are given below:

- `exp_act_win`: $$f_{act-win}(x,y)=\begin{cases}
        1, & \text{for } \sum_{i=1}^{m} x_i \ge \sum_{i=1}^{k} y_i\\
        0, & \text{otherwise}
        \end{cases}$$
- `exp_inh_win`: $$f_{inh-win}(x,y)=\begin{cases}
        1, & \text{for } \sum_{i=1}^{m} x_i \gt \sum_{i=1}^{k} y_i\\
        0, & \text{otherwise}
        \end{cases}$$

Note that: $f_{inh-win}(x,y) = \lnot f_{act-win}(y,x)$.

:::{.blue-box}
Note that every boolean function has an equivalent combinatorial expression (i.e. the variables are connected with logical operators) and I searched for an analytic formula for the two last threshold functions.
More more info, see this [math.stackexchange question](https://math.stackexchange.com/questions/3767774/identify-boolean-function-that-satisfies-some-constrains/).
:::
# Truth Density Data Analysis {-}

## Data {-}

:::{.blue-box}
*Truth Density (TD)* of a boolean equation/expression/function, given it's equivalent truth table, is the **number of rows that the function is active** divided by **the total number of rows** $(2^n)$.
:::

I created every possible truth table for up to $20$ variables (variables here means *regulators* for us) and calculated the `AND-NOT`, `OR-NOT`, `BalanceOp1`, `exp_act_win`, `exp_inh_win` boolean function results for every possible configuration of the number of activators and inhibitors that added up to the number of regulators.
Then, from the truth tables I calculated the **truth density** of each operator for each particular configuration.
For more details see the script [get_stats.R](get_stats.R)

:::{#tr-data}
See part of the data below:
:::
```{r load-data}
stats = readRDS(file = "data/td_stats.rds")

DT::datatable(data = stats,
  caption = htmltools::tags$caption("Truth Density Data", style="color:#dd4814; font-size: 18px"),
  options = list(pageLength = 6, scrollX = TRUE, order = list(list(1, "asc")))) %>% 
  formatRound(4:8, digits = 2)
```

## Truth Density formulas {-}

Here we prove the exact formulas for the truth densities in the case of the `AND-NOT` and `OR-NOT` link operator boolean functions.
For both propositions presented, we assume that $f$ is a boolean function $f(x,y):\{0,1\}^n \rightarrow \{0,1\}$, with a total of $n$ regulators/input variables.
We also assume that these regulators are uniquely separated to two distinct groups: $m$ **activators** (positive regulators) $x=\{x_i\}_{i=1}^{m}$ and $k$ **inhibitors** (negative regulators) $y=\{y_i\}_{i=1}^{k}$, with $n = m + k$.

```{proposition, name="AND-NOT Truth Density Formula", label='and-not-proof'}
When a boolean regulatory function $f$ has the form of an *AND-NOT* link-operator boolean function:
$f_{AND-NOT}(x,y) = \left(\bigvee_{i=1}^{m} x_i\right) \land \lnot \left(\bigvee_{i=1}^{k} y_i\right)$ with $m \ge 1$ activators and $k \ge 1$ inhibitors, its truth density is $TD_{AND-NOT}=\frac{2^m-1}{2^n} = \frac{1}{2^k}-\frac{1}{2^n}$.
```

```{proof}
Using the distributivity property and De Morgan's law we can re-write $f_{AND-NOT}$ in a DNF form as:
\begin{equation}
\begin{split}
f_{AND-NOT}(x,y) & = \left(\bigvee_{i=1}^{m} x_i\right) \land \lnot \left(\bigvee_{i=1}^{k} y_i\right) \\
                 & = \bigvee_{i=1}^{m} \left( x_i \land \lnot \left( \bigvee_{i=1}^{k} y_i \right) \right) \\ 
                 & = \bigvee_{i=1}^{m} (x_i \land \bigwedge_{j=1}^{k} \lnot y_j) \\ 
                 & = \bigvee_{i=1}^{m} (x_i \land \lnot y_1 \land ... \land \lnot y_k)
\end{split}
\end{equation}

To calculate the $TD_{AND-NOT}$, we need to find the number of rows of the $f_{AND-NOT}$ truth table that result in a *TRUE* output result and divide that by the total number of rows, which is $2^n$ ($n$ regulators/inputs).
Note that $f_{AND-NOT}$, written in it's equivalent DNF form, has exactly $m$ clauses.
Each clause has a unique *TRUE/FALSE* assignment of regulators that makes it *TRUE*.
This happens when the activator of the clause is *TRUE* and all of the inhibitors *FALSE*.
Since the condition for the inhibitors is the same regardless of the clause we are looking at and $f$ is expressed in a DNF form, the *TRUE* outcomes of the function $f$ are defined by all logical assignment combinations of the $m$ activators that have at least one of them being *TRUE* and all inhibitors assigned as *FALSE*.
There are a total of $2^m$ possible $TRUE/FALSE$ logical assignments of the $m$ activators (from all *FALSE* to all *TRUE*) and $f_{AND-NOT}$ becomes *TRUE* on all except one of them (i.e. when all activators are *FALSE*) and the corresponding $2^m-1$ truth table rows have all inhibitors assigned as *FALSE*.
Thus $TD_{AND-NOT}=\frac{2^m-1}{2^n}$.
```

<br>

```{proposition, name="OR-NOT Truth Density Formula", label='or-not-proof'}
When a boolean regulatory function $f$ has the form of an *OR-NOT* link-operator boolean function:
$f_{OR-NOT}(x,y) = \left(\bigvee_{i=1}^{m} x_i\right) \lor \lnot \left(\bigvee_{i=1}^{k} y_i\right)$ with $m \ge 1$ activators and $k \ge 1$ inhibitors, its truth density is $TD_{OR-NOT}=\frac{2^n-(2^k-1)}{2^n} = 1 + \frac{1}{2^n} - \frac{1}{2^m}$.
```

```{proof}
Using De Morgan’s law we can re-write $f_{OR-NOT}$ in a DNF form as:
\begin{equation}
\begin{split}
f_{OR-NOT}(x,y) & = \left(\bigvee_{i=1}^{m} x_i\right) \lor \lnot \left(\bigvee_{i=1}^{k} y_i\right) \\
                & = \left(\bigvee_{i=1}^{m} x_i\right) \lor \left(\bigwedge_{i=1}^{k} \lnot y_i\right) \\
                & = x_1 \lor x_2 \lor ... \lor x_m \lor (\lnot y_1 \land ... \land \lnot y_k)
\end{split}
\end{equation}

To calculate the $TD_{OR-NOT}$, it's easier to find the number of rows of the $f_{OR-NOT}$ truth table that result in a *FALSE* output result ($R_{false}$), subtract that number from the total number of rows ($2^n$) to get the rows that result in $f$ being *TRUE* and then divide by the total.
Thus $TD_{OR-NOT} = \frac{2^n-R_{false}}{2^n}$.
Note that $f_{OR-NOT}$, written in it's equivalent DNF form, has exactly $m+1$ clauses.
To make $f_{OR-NOT}$ *FALSE*, we first need to assign the $m$ activators as *FALSE* and then it all depends on the logical assignments of the inhibitors $y_i$ that are part of the last clause.
Out of all possible $2^k$ *TRUE/FALSE* logical assignments of the $k$ inhibitors (ranging from all *FALSE* to all *TRUE*) there is **only one that does not make** the last clause of $f_{OR-NOT}$ *FALSE* and that happens when all $k$ inhibitors are *FALSE*.
Thus, $R_{false}=2^k-1$ and $TD_{OR-NOT}=\frac{2^n-(2^k-1)}{2^n}$.
```

### Asymptotic behavior {-}

For a large number of regulators $n$, the truth densities of the `AND-NOT` and `OR-NOT` functions are as follows:

- `AND-NOT`: $$TD_{AND-NOT} = \frac{1}{2^k}-\frac{1}{2^n} \xrightarrow{n \text{ large}} \frac{1}{2^k} \xrightarrow{k \text{ large}}0$$
- `OR-NOT`:  $$TD_{OR-NOT} = 1 + \frac{1}{2^n} - \frac{1}{2^m} \xrightarrow{n \text{ large}} 1-\frac{1}{2^m} \xrightarrow{m \text{ large}} 1$$

:::{.green-box}
For large $n$, the $TD_{AND-NOT}$ depends only **on the number of inhibitors** and tends towards $0$ with increasing number of inhibitors.
Similarly, the $TD_{OR-NOT}$ depends only **on the number of activators** for large $n$ and tends towards $1$ with increasing number of activators.
:::

Also, again for large $n$, the extreme case of having a TD value equal to $1/2$ is a result of having **only one of the regulators being an inhibitor (resp. activator)** of the `AND-NOT` (resp. `OR-NOT`) equation.

### Validation {-}

We can use the [data above](#tr-data) to validate the [TD formulas](#truth-density-formulas) (up to $n=20$):
```{r validate-formulas, comment=""}
# Validate AND-NOT Truth Density formula
formula_td_and_not = stats %>% 
  mutate(formula_td_and_not = (2^num_act - 1)/(2^num_reg)) %>%
  pull(formula_td_and_not)

all(stats %>% pull(td_and_not) == formula_td_and_not)

# Validate OR-NOT Truth Density formula
formula_td_or_not = stats %>% 
  mutate(formula_td_or_not = (((2^num_act - 1) * (2^num_inh)) + 1)/(2^num_reg)) %>%
  pull(formula_td_or_not)

all(stats %>% pull(td_or_not) == formula_td_or_not)
```

## *AND-NOT* vs *OR-NOT* TD {-#stand-eq-bias}

Comparing the `AND-NOT` and `OR-NOT` truth densities across the number of regulators:
```{r fig-and-or-not, fig.align='center', dpi=300, cache=TRUE, fig.cap="AND-NOT vs OR-NOT Truth Densities across all possible activators and inhibitors combinations up to 20 regulators"}
# tidy up data
stats_and_or = tidyr::pivot_longer(data = stats, cols = c(td_and_not, td_or_not), 
  names_to = "lo", values_to = "td") %>%
  select(num_reg, lo, td) %>%
  mutate(lo = replace(x = lo, list = lo == "td_and_not", values = "AND-NOT")) %>%
  mutate(lo = replace(x = lo, list = lo == "td_or_not", values = "OR-NOT")) %>%
  rename(`Link Operator` = lo)

ggboxplot(data = stats_and_or, x = "num_reg", y = "td", 
  color = "Link Operator", palette = "Set1",
  title = "AND-NOT vs OR-NOT Truth Densities", 
  xlab = "Number of regulators", ylab = "Truth Density") +
  theme(plot.title = element_text(hjust = 0.5))
```

:::{.green-box}
- **The more regulators** there are, the more likely it is that the `AND-NOT` link operator in the boolean equation will result in an **inhibited** target and that the `OR-NOT` link operator in an **active** target.
- For $n>6$, the points outside the boxplots (with a truth density of $\frac{1}{2}, \frac{1}{4}, 1-\frac{1}{4},\frac{1}{8},1-\frac{1}{8},...$) correspond to the **long-term behavior** of the truth density formulas shown above where there is also **large imbalance between the number of activators and inhibitors**.
:::

We can also check the relation between TD and number of activators and inhibitors in each case.
The following two figures show us why **the number of inhibitors** are more decisive in the `AND-NOT` case:
```{r and-not-reg-plot, dpi=300, cache=TRUE, fig.show='hold', out.width='50%', fig.cap="AND-NOT TD vs Number of Activators and Inhibitors"}
ggscatter(data = stats %>% rename(`#Regulators` = num_reg), x = "num_inh", 
  y = "td_and_not", color = "#Regulators",
  ylab = "Truth Density", xlab = "Number of Inhibitors", 
  title = "AND-NOT TD vs Number of Inhibitors") + 
  theme(plot.title = element_text(hjust = 0.5))

ggscatter(data = stats %>% rename(`#Regulators` = num_reg), x = "num_act", 
  y = "td_and_not", color = "#Regulators",
  ylab = "Truth Density", xlab = "Number of Activators", 
  title = "AND-NOT TD vs Number of Activators") + 
  theme(plot.title = element_text(hjust = 0.5))
```

In the `OR-NOT` case the number of activators is more important:
```{r or-not-reg-plot, dpi=300, cache=TRUE, fig.show='hold', out.width='50%', fig.cap="OR-NOT TD vs Number of Activators and Inhibitors"}
ggscatter(data = stats %>% rename(`#Regulators` = num_reg), 
  x = "num_inh", y = "td_or_not", color = "#Regulators",
  ylab = "Truth Density", xlab = "Number of Inhibitors", 
  title = "OR-NOT TD vs Number of Inhibitors") + 
  theme(plot.title = element_text(hjust = 0.5))

ggscatter(data = stats %>% rename(`#Regulators` = num_reg), 
  x = "num_act", y = "td_or_not", color = "#Regulators",
  ylab = "Truth Density", xlab = "Number of Activators", 
  title = "OR-NOT TD vs Number of Activators") + 
  theme(plot.title = element_text(hjust = 0.5))
```

## *BalanceOp1* TD {-}

If we add the `BalanceOp1` TD results to the first plot we have:
```{r fig-and-or-not-balanceop, fig.align='center', dpi=300, cache=TRUE, fig.cap="AND-NOT vs OR-NOT vs BalanceOp1 Truth Densities across all possible activators and inhibitors combinations up to 20 regulators"}
# tidy up data
stats_and_or_balance = tidyr::pivot_longer(data = stats, cols = c(td_and_not, td_or_not, td_balance_op), 
  names_to = "lo", values_to = "td") %>%
  select(num_reg, lo, td) %>%
  mutate(lo = replace(x = lo, list = lo == "td_and_not", values = "AND-NOT")) %>%
  mutate(lo = replace(x = lo, list = lo == "td_or_not", values = "OR-NOT")) %>%
  mutate(lo = replace(x = lo, list = lo == "td_balance_op", values = "BalanceOp1")) %>%
  rename(`Link Operator` = lo)

ggboxplot(data = stats_and_or_balance, x = "num_reg", y = "td", 
  color = "Link Operator", palette = "Set1",
  title = "AND-NOT vs OR-NOT vs BalanceOp1 Truth Densities", 
  xlab = "Number of regulators", ylab = "Truth Density") +
  theme(plot.title = element_text(hjust = 0.5))
```

:::{.green-box}
- The `BalanceOp1` TD values are closer to the TD values of the `OR-NOT` formula compared to the `AND-NOT` one.
- The `BalanceOp1` is less *biased* compared to the `OR-NOT` link operator, but still for large $n$ (regulators) it practically **makes the target activated**.
:::

As we can see in the following two figures, the `BalanceOp1` shows a more balanced dependency between the number of activators and inhibitors:
```{r balanceOp1-reg-plot, dpi=300, cache=TRUE, fig.show='hold', out.width='50%', fig.cap="BalanceOp1 TD vs Number of Activators and Inhibitors"}
ggscatter(data = stats %>% rename(`#Regulators` = num_reg), x = "num_inh", 
  y = "td_balance_op", color = "#Regulators",
  ylab = "Truth Density", xlab = "Number of Inhibitors", 
  title = "BalanceOp1 TD vs Number of Inhibitors") + 
  theme(plot.title = element_text(hjust = 0.5))

ggscatter(data = stats %>% rename(`#Regulators` = num_reg), x = "num_act", 
  y = "td_balance_op", color = "#Regulators",
  ylab = "Truth Density", xlab = "Number of Activators", 
  title = "BalanceOp1 TD vs Number of Activators") + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Threshold Functions TD {-}

In contrast, if we check the truth density of the $f_{act-win}(x,y)$ and $f_{inh-win}(x,y)$ boolean functions we have:
```{r fig-two-bool-formulas, fig.align='center', dpi=300, cache=TRUE, fig.cap="Truth Desities of two robust boolean formulas across all possible activators and inhibitors combinations up to 20 regulators"}
# tidy up data
stats_functions = tidyr::pivot_longer(data = stats, cols = c(td_exp_act, td_exp_inh), 
  names_to = "fun", values_to = "td") %>%
  select(num_reg, fun, td) %>%
  mutate(fun = replace(x = fun, list = fun == "td_exp_act", values = "Activators Win")) %>%
  mutate(fun = replace(x = fun, list = fun == "td_exp_inh", values = "Inhibitors Win")) %>%
  rename(`Equation Formula` = fun)

ggboxplot(data = stats_functions, x = "num_reg", y = "td",
  color = "Equation Formula", palette = "lancet",
  title = latex2exp::TeX("Truth Densities of $f_{act-win}(x,y)$ and $f_{inh-win}(x,y)$"),
  xlab = "Number of regulators", ylab = "Truth Density") +
  theme(plot.title = element_text(hjust = 0.5))
```

:::{.green-box}
- Both boolean functions show a **large variance of truth densities** irrespective of the number of regulators, since the values inside the boxplots represent the middle $50\%$ of the data and span across the $(0,1)$ range.
- The median values seem to converge to $0.5$ for both formulas.
- The median value of truth density for the $f_{act-win}(x,y)$ is always larger than the $f_{inh-win}(x,y)$ (as expected).
:::

## TD Data Distance {-}

We check how close are the truth density values of the different proposed BBRs, also compared to the **proportion of activators**, e.g. if a BBR has 1 activator and 5 inhibitors (resp. 5 activators and 1 inhibitor) I would expect the boolean regulatory function's output to be statistically more inhibited (resp. activated).
We find the *euclidean distance* between the different truth density values and show them in a table and dendrogram format:

```{r td-values-distance}
act_prop = stats %>% mutate(act_prop = num_act/num_reg) %>% pull(act_prop)
td_and_not = stats %>% pull(td_and_not)
td_or_not = stats %>% pull(td_or_not)
td_balance_op = stats %>% pull(td_balance_op)
td_exp_act = stats %>% pull(td_exp_act)
td_exp_inh = stats %>% pull(td_exp_inh)

d = dist(x = rbind(act_prop, td_and_not, td_or_not, td_balance_op, td_exp_act, td_exp_inh),
         method = "euclidean")
```

```{r dist-table}
# color `act_prop` column
breaks = quantile(unname(as.matrix(d)[, "act_prop"]), probs = seq(.05, .95, .05), na.rm = TRUE)
col = round(seq(255, 40, length.out = length(breaks) + 1), 0) %>%
  {paste0("rgb(255,", ., ",", ., ")")} # red

caption.title = "Euclidean Distances between vectors of truth density values (Symmetric)"
DT::datatable(data = d %>% as.matrix(), options = list(dom = "t", scrollX = TRUE),
  caption = htmltools::tags$caption(caption.title, style="color:#dd4814; font-size: 18px")) %>% 
  formatRound(1:6, digits = 3) %>%
  formatStyle(columns = c("act_prop"), backgroundColor = styleInterval(breaks, col))
```

```{r dist-dendogram, fig.height=7, fig.align='center'}
plot(hclust(dist(d)), main = "Distance Dendogram of Thruth Densities",
  ylab = "Euclidean Distance", sub = "", xlab = "")
```

:::{.green-box}
- The **threshold functions** have truth densities values that are **closer to the proportion of activators** for a varying number of regulators, compared to the `AND-NOT` and `OR-NOT` formulas.
As such they might represent more realistic candidates for regulatory functions from a statistical point of view.
- The TD values of `OR-NOT` and `BalanceOp1` are in general very close (as we've also seen in a previous figure)
:::

## Correlation {-}

We will now check the *correlation* between each pair of operators/proposed functions, as well as the number of regulators, inhibitors and activators:
```{r cor-plot, fig.align='center', dpi=300, cache=TRUE, fig.cap="Correlation Matrix of Truth Densities and number of regulators"}
M = cor(stats, method = "kendall")
res = corrplot::cor.mtest(stats, method = "kendall")
corrplot::corrplot(corr = M, type = "upper", p.mat = res$p, sig.level = c(.001, .01, .05), 
  pch.cex = 1, pch.col = "white", insig = "label_sig", tl.col = "black", tl.srt = 45)
```

:::{.green-box}
- The two functions results $f_{act-win}(x,y), f_{inh-win}(x,y)$ are highly correlated as expected
- Lower `AND-NOT` TD values highly correlate with *higher* number of inhibitors
- Higher `OR-NOT` TD values highly correlate with *higher* number of activators
:::


# CASCADE 1.0 Data Analysis {-}

## Data {-}

Using [abmlog](https://github.com/druglogics/abmlog) we generated all $2^{23} = 8388608$ possible link operator mutated models for the CASCADE 1.0 topology ($23$ nodes have at least one regulator from each category, i.e. an activator and an inhibitor).
The models are stored in both `.gitsbe` and `.bnet` files in the Zenodo dataset [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4022783.svg)](https://doi.org/10.5281/zenodo.4022783) (the `gitsbe` files include also the fixpoint attractors).

:::{.blue-box}
The dataset includes models with $0,1$ or $2$ stable states.
We used the [get_ss_data.R](https://github.com/bblodfon/brf-bias/blob/main/scripts/get_ss_data.R) script to get the **1 stable state data** from the models and it's this data we are going to analyze in the next section.
:::

## Parameterization vs Activity {-}

We calculate the `node_stats` object using the [get_node_stats.R](https://github.com/bblodfon/brf-bias/blob/main/scripts/get_node_stats.R) script.
This object includes the **agreement statistics information** for each link-operator node (i.e. one that is targeted by both activators and inhibitors).

Load the `node_stats`:
```{r load_node_stats}
node_stats = readRDS(file = "data/node_stats.rds")
```

:::{.note}
We are interested in two variables of interest:

- **Parameterization** of a link operator node: `AND-NOT` (0) vs `OR-NOT` (1)
- **Stable State** of a node: *inhibited* (0) vs *active* (1)

There exist are 4 different possibilities related to 2 cases:

1. `0-0`, `1-1` => parameterization and stable state **match** (e.g. node was parameterized with `AND-NOT` and it's state was inhibited or it had `OR-NOT` and its state was active)
2. `1-0`, `0-1` => parameterization and stable state **differ** (e.g. node had `OR-NOT` and its state was inhibited, or `AND-NOT` and it's state was active)
:::

In the next Figure we show the **total observed proportionate agreement or percent agreement** for each node, which is the number of models for which **parameterization and stable state matched (case 1 above) divided by the total amount of models**:
```{r ss-lo-agreement-prop, fig.align='center', dpi=300, cache=TRUE, fig.cap="Parameterization and Stable State activity agreement"}
node_stats %>% mutate(node = forcats::fct_reorder(node, desc(num_reg))) %>% 
  ggplot(aes(x = node, y = obs_prop_agreement, fill = as.factor(num_reg))) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels=scales::percent) +
    labs(title = "Agreement between Link Operator Parameterization and Stable State Activity", x = "Target Nodes with both activating and inhibiting regulators", y = "Percent Agreement") +
    theme_classic() + theme(axis.text.x = element_text(angle = 90)) +
    scale_fill_brewer(guide = guide_legend(reverse=TRUE, title = "#Regulators"), palette = "Set1") +
    geom_hline(yintercept = 0.5, linetype = 'dashed')
```

:::{.green-box}
The total barplot area covered (i.e. the **total percent agreement score** so to speak) is **`r node_stats %>% summarise(100 * sum(obs_prop_agreement)/n()) %>% pull()`%**.

The above score means that the is a higher probability than chance to assign a node the `AND-NOT` (resp. `OR-NOT`) link operator in its respective boolean equation and that node at the same time having an inhibited (resp. activated) stable state of $0$ (.resp $1$) in any CASCADE 1.0 link operator parameterized model.
**This suggests that the corresponding boolean formula used is biased**.
:::

In the next figure, where we have **separated the nodes to groups based on the number of regulators**, we observe that even for a small number of regulators, the **median percent agreement score is higher than** $0.5$:
```{r ss-lo-agreement-prop-grouped, fig.align='center', warning=FALSE, dpi=300, cache=TRUE, fig.cap="Parameterization and Stable State activity agreement. CASCADE 1.0 link-operator nodes are grouped based on their respective number of regulators"}
node_stats %>% 
  mutate(num_reg = as.factor(num_reg)) %>%
  ggplot(aes(x = num_reg, y = obs_prop_agreement, fill = num_reg)) + 
    geom_boxplot(show.legend = FALSE) +
    geom_jitter(shape = 19, position = position_jitter(0.2), show.legend = FALSE) +
    scale_y_continuous(labels = scales::percent, limits = c(0,1)) +
    labs(title = "Agreement (parameterization vs stable state activity)", x = "Number of Regulators", y = "Percent Agreement") +
    geom_hline(yintercept = 0.5, linetype = 'dashed', color = "red") +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(size = 15))
```

Next, we calculate per node, the proportion of link operator assignments that retained their expected (i.e. keeping the same digit) stable state activity (e.g. the proportion of models corresponding to the cases `0-0`/(`0-0` + `0-1`) for the `AND-NOT` link operator - and `1-1`/(`1-1` + `1-o`) for `OR-NOT`):
```{r ss-comp-agreement-props, fig.align='center', dpi=300, cache=TRUE, fig.cap="Parameterization and Stable State activity agreement 2"}
node_stats %>% 
  mutate(and_not_0ss_prop = and_not_0ss_agreement/(and_not_0ss_agreement + and_not_1ss_disagreement)) %>% 
  mutate(or_not_1ss_prop  = or_not_1ss_agreement/(or_not_1ss_agreement + or_not_0ss_disagreement)) %>%
  select(node, num_reg, and_not_0ss_prop, or_not_1ss_prop, active_prop) %>%
  rename(`AND-NOT` = and_not_0ss_prop, `OR-NOT` = or_not_1ss_prop) %>%
  mutate(node = forcats::fct_reorder(node, desc(num_reg))) %>%
  pivot_longer(cols = c(`AND-NOT`, `OR-NOT`)) %>%
  ggplot(aes(x = node, y = value, fill = name)) +
    geom_bar(position = "dodge", stat = "identity") +
    scale_y_continuous(labels=scales::percent) +
    labs(title = "Link Operator Parameterization Agreement with Stable State Activity", 
      x = "Target Nodes with both activating and inhibiting regulators", 
      y = "Percent Agreement") +
    theme_classic() + theme(axis.text.x = element_text(angle = 90)) + 
    scale_fill_brewer(guide = guide_legend(title = "Link Operator"), palette = "Set1") + 
    geom_line(aes(y = active_prop, color = active_prop), group = 1, size = 1.2) +
    scale_color_gradient(labels=scales::percent, low="grey", high="green", 
      name = "%Models:active node", limits = c(0,1)) + 
    theme(legend.title = element_text(size = 10))
```

:::{.green-box}
- Higher proportional activity for a node correlates with higher `OR-NOT`-activated state agreement.
- `LRP_f` has 4 activators and 1 inhibitor and from the previous [TD data table](#tr-data) we have that: $TD_{AND-NOT,4+1}=0.469$, $TD_{OR-NOT,4+1}=0.969$, numbers which correspond really well with the percent agreement scores found across all the CASCADE 1.0 models.
- `TSC_f` and `mTORC2_c` are always found inhibited and thus the agreement with the `AND-NOT`-inhibited state is perfect and the `OR-NOT`-activated state agreement zero.
- `TSC_f` has 1 activator and 4 inhibitors, which corresponds well to it's total inhibition profile in all the models (with significantly more inhibitors, there is a higher probability of the target node being inhibited).
The TD values are $TD_{AND-NOT,1+4}=0.03$, $TD_{OR-NOT,1+4}=0.53$ and so the percent agreement between the `AND-NOT` parameterization and the resulting $0$ stable state activity is justified, but for the `OR-NOT` cases we would expect around half of them to be in agreement (have a value of $1$ in the stable state) - which was not the case (all of them had $0$).
Probably the network dynamical configuration may have something to do with that.
:::

```{r lo-stats-table}
caption.title = "Link Operator Statistics (CASCADE 1.0)"
DT::datatable(data = node_stats %>% select(node, num_reg, num_act, num_inh), 
  caption = htmltools::tags$caption(caption.title, style="color:#dd4814; font-size: 18px"),
  options = list(order = list(list(2, "desc")))) %>% 
  formatRound(5:6, digits = 3)
```

# CASCADE 2.0 Data Analysis {-}

:::{.blue-box}
We will perform the same analysis as in the previous section, only now for a **randomly selected sample of models from the CASCADE 2.0 topology**.

CASCADE 2.0 represents a larger topology/network with nodes with more than $5$ regulators and as such we expect to see even more agreement between stable state activity and link operator assignment for these nodes (which will be a proof-of-concept for the link operator bias).
:::

## Data {-}

:::{.note}
The dataset used was generated for [another analysis](https://bblodfon.github.io/gitsbe-model-analysis/cascade/random-model-ss/main.html) and we are going to use part of it, i.e. the **models that had 1 stable state** (see [get_node_stats_cascade_2.R](https://github.com/bblodfon/brf-bias/blob/main/scripts/get_node_stats_cascade_2.R) script).
The dataset is stored in Zenodo [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3932382.svg)](https://doi.org/10.5281/zenodo.3932382)
:::

Load the CASCADE 2.0 `node_stats_cascade2`:
```{r load-node-stats-cascade2}
node_stats_cascade2 = readRDS(file = "data/node_stats_cascade2.rds")
```

## Parameterization vs Activity {-}

The next Figure shows the **total percent agreement** for each link operator node in CASCADE 2.0 (a total of $52$ nodes), which is the number of models for which **parameterization and stable state matched** divided by the total amount of models ($20672$):
```{r ss-lo-prop-aggreement-cascade2, fig.align='center', dpi=300, cache=TRUE, fig.cap='Parameterization and Stable State activity agreement (CASCADE 2.0)'}
node_stats_cascade2 %>% mutate(node = forcats::fct_reorder(node, desc(num_reg))) %>% 
  ggplot(aes(x = node, y = obs_prop_agreement, fill = as.factor(num_reg))) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels=scales::percent) +
    labs(title = "Agreement between Link Operator Parameterization and Stable State Activity", x = "Target Nodes with both activating and inhibiting regulators", y = "Percent Agreement") +
    theme_classic() + theme(axis.text.x = element_text(angle = 90)) +
    scale_fill_brewer(guide = guide_legend(reverse=TRUE, title = "#Regulators"), palette = "Spectral") +
    geom_hline(yintercept = 0.5, linetype = 'dashed')
```
:::{.green-box}
The total barplot area covered (i.e. the **total agreement score** so to speak) is **`r node_stats_cascade2 %>% summarise(100 * sum(obs_prop_agreement)/n()) %>% pull()`%**.

The nodes with number of regulators $>5$ have always a percent agreement $\geq 70\%$ between stable state activity and link operator parameterization.
The above results provide evidence that the statistics-based conclusion we reached in a [previous section](#stand-eq-bias) is correct, i.e. that the **standardized boolean formula is biased for larger number of regulators**.
:::

In the next two figures, we group the nodes to $3$ groups, based on the number of regulators they have.
In order to account for the uncertainty factors involved when comparing stable state activity values and corresponding link-operator parameterization, we additionally try a more *robust* statistic to measure the level of agreement, namely the **Cohen's kappa coefficient** ($\kappa$):
```{r ss-lo-prop-aggreement-cascade2-grouped, fig.align='center', fig.show='hold', out.width='50%', dpi=300, cache=TRUE, fig.cap='Parameterization and Stable State activity agreement. CASCADE 2.0 link-operator nodes are grouped based on their respective number of regulators'}
node_stats_cascade2 %>% mutate(reg_group = 
    factor(case_when(num_reg >= 6 ~ ">5", num_reg <= 3 ~ "2-3", TRUE ~ "4-5"), levels = c("2-3", "4-5", ">5"))) %>%
  ggplot(aes(x = reg_group, y = obs_prop_agreement, fill = reg_group)) + 
    geom_boxplot(show.legend = FALSE) +
    geom_jitter(shape = 19, position = position_jitter(0.2), show.legend = FALSE) +
    scale_y_continuous(labels = scales::percent, limits = c(0,1)) +
    labs(title = "Agreement (parameterization vs stable state activity)", x = "Number of Regulators", y = "Percent Agreement") +
    geom_hline(yintercept = 0.5, linetype = 'dashed', color = "red") +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(size = 15))

node_stats_cascade2 %>% mutate(reg_group = 
    factor(case_when(num_reg >= 6 ~ ">5", num_reg <= 3 ~ "2-3", TRUE ~ "4-5"), levels = c("2-3", "4-5", ">5"))) %>%
  ggplot(aes(x = reg_group, y = cohen_k, fill = reg_group)) + 
    geom_boxplot(show.legend = FALSE) +
    geom_jitter(shape = 19, position = position_jitter(0.2), show.legend = FALSE) +
    ylim(c(0,1)) +
    labs(title = "Cohen's k (parameterization vs stable state activity)", x = "Number of Regulators", y = latex2exp::TeX("$\\kappa$")) +
    geom_hline(yintercept = 0.6, linetype = 'dashed', color = "red") +
    geom_text(aes(x = 3.4, y = 0.55, label="k = 0.6")) + 
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(size = 15))
```
:::{.green-box}
The larger the number of regulators, the higher the percent agreement and Cohen's $\kappa$.
:::

:::{.note}
Though **Cohen's kappa** $\kappa$ might not be a proper agreement measurement for this dataset, we know that it is more sophisticated than the simple percent agreement calculation, as it takes into account the possibility of the agreement occurring by chance.

In the figure above we have drawn a horizontal line to indicate the value of $\kappa$ that corresponds to a *moderate* or *substantial* strength of agreement [@Landis1977; @McHugh2012a]
:::

# R session info {-}

```{r session info, comment=""}
xfun::session_info()
```

# References {-}
